---
title: "Reproducible Research Assignment"
output: html_document
---

We first download the data file, and then load it into R. The Date column in converted into a Date type.

```{r echo=TRUE}
data <- read.csv("repdata-data-activity\\activity.csv")
data$date <- as.Date(data$date, "%Y-%m-%d")

```

To calculate the number of steps per day, I used the plyr package.
```{r echo=TRUE}
library(plyr)
databyday<-ddply(data, .(date), 
      summarize, 
      stepsbyday = sum(steps, na.rm=TRUE))


```
a histogram of the total number of steps taken each day
``` {r echo=TRUE}
hist(databyday$steps, breaks=20,main="Frequency of Number of Steps", xlab="Number of steps")

```
the mean and median of the total number of steps taken per day
```{r echo=TRUE}
meanbyday<-mean(databyday$steps)
medianbyday<-median(databyday$steps)
meanbyday
medianbyday
```

Next, we make a time series plot, to show the average number of steps take across all days

```{r, echo=FALSE}
databyinterval<- ddply(data, .(interval),summarize, stepsbyinterval =mean(steps, na.rm=TRUE))

plot(databyinterval$interval, databyinterval$stepsbyinterval, type="l", xlab="Interval", ylab="average steps", main="Average steps by interval")

```
The 5 min interval, which contains the average maxiumum number of steps is
``` {r echo=TRUE}
databyinterval[which.max(databyinterval$stepsbyinterval),2]
```

To view the total number of missing values, we can view the summary of the initial dataset
```{r}
summary(data)

```

There are 2304 NA values for steps.

To replace the missing values, I am substituting the values with the average value for that interval
```{r}
mergedData <- merge(data, databyinterval, by="interval")


for(i in 1:nrow(mergedData)) {
  if(is.na(mergedData[i,"steps"]))
    mergedData[i,"steps"]<-mergedData[i,"stepsbyinterval"]
    
}

completeData<-ddply(mergedData, .(date), 
      summarize, 
      SumOfStepsbyDay = sum(steps))
hist(completeData$SumOfStepsbyDay, breaks=20,main="Frequency of Number of Steps without NA", xlab="Number of steps")
mean(completeData$SumOfStepsbyDay)
median(completeData$SumOfStepsbyDay)

```

The difference in the 2 histograms is that in the first one, we had a high number of 0s. However,  once those were replaced by the mean values of the respective intervals, we now see a jump in the frequency of the the steps that are close the mean values or the center of the histogram.

Creating a new factor variable:weekday
```{r}
mergedData$day<- "weekday"
for(i in 1:nrow(mergedData)) {
  if(weekdays(mergedData[i,"date"]) =="Saturday" || weekdays(mergedData[i,"date"]) =="Sunday")
              mergedData[i,"day"]<-"weekend"
      
}
mergedData$day<-as.factor(mergedData$day)

```

Generating the panel plot
```{r echo=TRUE}
library(ggplot2)
avgsteps<-ddply(mergedData,.(interval,day), summarize, avg_steps=mean(steps))
sp<-ggplot(data=avgsteps, aes(x=interval, y=avg_steps, group=1)) +
    geom_line()
sp + facet_grid(day ~ .)
```

